Dissecting the relationship between protein structure and sequence variation

### 2015-05-16 manuscript/voronoi/mnv.tex
MN updated

\vfill
%\Title{Predicting Sequence Variability from Voronoi Tessellation of Proteins}
\Title{Structural Determinants of Sequence Evolution in Enzymatic proteins}
\vfill
\Author{Amir Shahmoradi$^{1}$, Claus O. Wilke$^2$}
\Address{\affiliation}
\vfill
\begin{Abstract}
    What are the best structural predictors of protein's sequence evolution? A number of site-specific structural properties have been proposed over the past decade to answer this question. The majority of these quantities however, depend on the set of atomic coordinates used to represent individual sites in proteins, and often involve one or more number of adjustable parameters in their definition. A number of studies have already demonstrated that the choice of $C_\alpha$ atomic coordinates may not be an optimal representation of the protein's 3-dimensional structure, in particular for the calculation of site-specific quantities such as Weighted Contact Number.  Expanding on these studies, here we propose a new set of parameter-free structural variables derived from the Voronoi tessellation of protein structure which perform equally well or better than virtually all previously-considered structural quantities in predicting protein sequence evolution. We further show that the ideal representation of the 3-dimensional structure of proteins is the set of geometric average coordinates of atoms in the side chains of individual amino acids versus the common choice of backbone $C_\alpha$ coordinates. Long-range interactions between individual amino acids are found to have a minor but significant contribution to the evolutionary patterns of protein sequence.
    % OLD abstract:
    %A dataset of 209 monomeric enzymes is considered here to carry out a comprehensive search for the potential structural determinants of the site-specific evolution of protein sequence. Based on Voronoi tessellation of protein structures, we define new site-specific structural properties that are on average capable of describing up to $30\%$ of sequence variability observed in the dataset. We show that the Voronoi cell area and volume outperform other structural proxy measures of site-specific sequence variability previously considered in the literature, such as the Relative Solvent Accessibility, Contact Number (CN), and measures of local flexibility such as the Debye-Waller factor.  Using a variety of atomic coordinates in the definition and calculation of structural properties, we show that the best representative set of coordinates for individual sites in proteins is the average of the side-chain atomic coordinates. This choice of coordinates for the definition of structural properties results in the best predictions of site-specific sequence variability. Example structural properties that show significant improvement in this regard include those derived from Voronoi tessellation, contact number and representative B-factors for individual sites in proteins. On average, the choice of geometrical centers of side-chains vs. the commonly chosen coordinates of the backbone atom $C_\alpha$ can result in $0.07$ to $0.12$ improvements in sequence-structure correlation strengths. We finally argue and show that there is no uniquely-defined best kernel for the calculation of the Weighted Contact Number, which is commonly defined as the sum of the inverse-square of the reciprocal distances between pairs of sites in proteins. By contrast, other definitions perform equally well or better.  This finding highlights the diverse energy landscapes of proteins and the fact that no single potential-of-mean-force can uniquely describe all interactions between individual sites in proteins.
\end{Abstract}
\vfill
\vfill
\end{titlepage}
\def\thefootnote{\fnsymbol{footnote}}
\setcounter{footnote}{0}
%

>>>>

\vfill
%\Title{Predicting Sequence Variability from Voronoi Tessellation of Proteins}
\Title{Dissecting the relationship between protein structure and sequence variation}
\vfill
\Author{Amir Shahmoradi$^{1}$, Claus O. Wilke$^2$}
\Address{\affiliation}
\vfill
\begin{Abstract}
    What are the best structural predictors of protein's sequence evolution? A number of site-specific structural properties have been proposed over the past decade to answer this question. Most importantly, the role of local packing density has been highlighted recently and shown to be the dominant factor in the observed patterns of site-specific sequence variability of proteins. A wide variety of local packing density measures has been introduced in the literature such as the Weighted Contact Number which represents by definition the combined effects of local packing density and long-range amino acid interactions on sequence variability. Here we propose a methodology to segregate and quantify the extent to which long range interactions determine the general patterns of sequence variability in proteins, independently of local packing density. We use Voronoi partitioning of protein's 3-dimensional structure to obtain a parameter-free measure of local packing cell, defined as the inverse of the volume of the Voronoi cells corresponding to each site in protein.

    The majority of these quantities however depend on the set of atomic coordinates used to represent individual sites in proteins, and often involve one or more number of adjustable parameters in their definition. A number of studies have already demonstrated that the choice of $C_\alpha$ atomic coordinates may not be an optimal representation of the protein's 3-dimensional structure, in particular for the calculation of site-specific quantities such as Weighted Contact Number.  Expanding on these studies, here we propose a new set of parameter-free structural variables derived from the Voronoi tessellation of protein structure which perform equally well or better than virtually all previously-considered structural quantities in predicting protein sequence evolution. We further show that the ideal representation of the 3-dimensional structure of proteins is the set of geometric average coordinates of atoms in the side chains of individual amino acids versus the common choice of backbone $C_\alpha$ coordinates. Long-range interactions between individual amino acids are found to have a minor but significant contribution to the evolutionary patterns of protein sequence.
    % OLD abstract:
    %A dataset of 209 monomeric enzymes is considered here to carry out a comprehensive search for the potential structural determinants of the site-specific evolution of protein sequence. Based on Voronoi tessellation of protein structures, we define new site-specific structural properties that are on average capable of describing up to $30\%$ of sequence variability observed in the dataset. We show that the Voronoi cell area and volume outperform other structural proxy measures of site-specific sequence variability previously considered in the literature, such as the Relative Solvent Accessibility, Contact Number (CN), and measures of local flexibility such as the Debye-Waller factor.  Using a variety of atomic coordinates in the definition and calculation of structural properties, we show that the best representative set of coordinates for individual sites in proteins is the average of the side-chain atomic coordinates. This choice of coordinates for the definition of structural properties results in the best predictions of site-specific sequence variability. Example structural properties that show significant improvement in this regard include those derived from Voronoi tessellation, contact number and representative B-factors for individual sites in proteins. On average, the choice of geometrical centers of side-chains vs. the commonly chosen coordinates of the backbone atom $C_\alpha$ can result in $0.07$ to $0.12$ improvements in sequence-structure correlation strengths. We finally argue and show that there is no uniquely-defined best kernel for the calculation of the Weighted Contact Number, which is commonly defined as the sum of the inverse-square of the reciprocal distances between pairs of sites in proteins. By contrast, other definitions perform equally well or better.  This finding highlights the diverse energy landscapes of proteins and the fact that no single potential-of-mean-force can uniquely describe all interactions between individual sites in proteins.
\end{Abstract}
\vfill
\vfill
\end{titlepage}
\def\thefootnote{\fnsymbol{footnote}}
\setcounter{footnote}{0}
%
